---
title: "Is denser population area has more crime than others?"
author: "Thu Dong"
format: pdf
editor: visual
abstract: "This is my abstract."
---

```{r}
#install.packages("kableExtra")
#install.packages( "psych")
```

```{r}
#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(psych)
```

```{r}
neighborhood_crime_clean <-
  read_csv(
    "neighborhood_crime_2023_clean.csv",
    show_col_types = FALSE
  )
```

## Table of Content

## 1. Introduction

Toronto is considered one of the safest cities in Canada and North America with homicide rates of approximately 1.95 per 100,000 people in 2020 (Armstrong and Jaffray 2021). However, it was reported that there was a 14% increase in the overall crime rate in 2022 and it will continue to increase in 2023 (Rosenbaum 2023). It is believed that more populated areas tend to have higher crime incidents. This paper attempts to investigate whether areas with higher populations experience higher levels of crime occurrences.  

There have been many studies in criminology that discuss the relationship between population size and crime level. One of the first works in criminology is research conducted across France in 1825 by Balbi and Guerry which reveals that the area with a higher population has more crime occur than others. However, it was later discussed that these 2 relationships are not linear and might be a correlation-based conclusion. (Friendly 2007). On the other hand, some other research in criminology disproved the theory that crime increases with population. (Chamlin and Cochran 2004). This research argued that there is no direct association between population and crime rate and other factors that are related to population growth such as quality of life have a stronger correlation to the volume of crime than population. Most of the research has used crime rates to measure the intensity of crime in an area, standardize a measure, and manage risk assessments. This paper will be using crime volume or the number of crimes to assess the magnitude of crime level. Using crime volume also helps policymakers to understand the safety state of each area to give efficient resource allocation. 

To discuss the relationship between population density and crime volume, this paper will present in the following section: Data, Results, Discussion, and Conclusion. In the Data section, I will discuss the nature of the data obtained through the City of Toronto\'s OpenDataToronto Library and the steps I took to clean the data. The Result section will highlight some trends found in the data. Additionally, the Discussion Section will explore the relationship between the level of crime occurrence and population density. Finally, The conclusion section will summarize the main findings of the paper. 

## 2. Data

The data used in this paper was obtained through the City of Toronto\'s OpenDataToronto Library. The data set is called \'Neighborhood Crime Rate\' (\"Open Data Dataset\" 2024). Data was collected and analyzed using the statistical programming software R (R Core Team 2023), with additional support from tidyverse (Wickham et al. 2019), ggplot2 (Wickham 2016), dplyr (Wickham et al. 2023), readr (Wickham, Hester, and Bryan 2023), **( add all package used)**

### **2.1.Crime volume in 2023.** 

The original \"Neighborhood Crime Rate\" Data set includes \_\_\_\_ Variables and \_\_\_ observations. It records the number of Assaults, Auto Theft, Break and Entering, Robbery, Theft Over, Homicide, and Shooting from the year 2014 to the year 2023, as well as the crime rate of each crime in each year. 

To get the data for crime volume in 2023, I selected 8 columns: Assaults, Auto Theft, Breaking and Entering, Robbery, Theft Over, Homicide, and Shooting in 2023 since it is the closest data recorded. Additionally, all the \"NA\" values were set to 0 to allow for mathematical operation on the data.  Then, I summed the number of each crime to get the total crime of each area in the year 2023. Finally, I renamed all the columns needed to allow for clarity.

### **2.2. Population in 2023.**

Data for area in Toronto and population in 2023 was available in the \" Neighborhood Crime rate\" dataset. Therefore, the name of the area and the population in 2023 were added to the clean data set. Table 1 shows the first 5 rows of the data set with area name, population, and the number of crime occurrences in 2023. 

 

```{r}

neighborhood_crime <- select(neighborhood_crime_clean, AREA_NAME, POPULATION_2023, TOTAL_CRIME_2023)

write_csv(neighborhood_crime, "neighborhood_crime_analysis.csv")

kable(head(neighborhood_crime,5)) %>%
  kable_styling(full_width = FALSE)

```

## 3. Result

### **3.1. Toronto Area and the population in 2023.**

There are 158 areas divided in Toronto. Out of the 158 areas in Toronto,  Beechborough-Greenbrook is the part with the lowest population of 7,057 and West Humber-Clairville is the area with the highest population at 36,388. Table 2 shows the 10 most populated areas in Toronto. The second most populated area is Mount Olive-Silverstone-Jamestown with 35,386 people but the third highest populated area is the Annex with 34502 people. Other area that are highly populated are Glenfield-Jane Heights, Harbourfront-CityPlace, Agincourt North, etc... 

```{r}
#Summary table
summary_result <- describe(neighborhood_crime$POPULATION_2023)
#summary_result

#area with lowest population
min_populated_area <- neighborhood_crime$AREA_NAME[which.min(neighborhood_crime$POPULATION_2023)]

sorted_population_table <- neighborhood_crime_clean %>%
  arrange(desc(POPULATION_2023)) %>%
  slice_head(n = 10)


sorted_population_table |>
  summarise(population = POPULATION_2023,
            .by = AREA_NAME) |>
  kable()

```

### **3.2. Toronto Area and crime volume.** 

Table 3 shows 10 areas with the highest level of criminal incidents in Toronto. Out of the 158 areas in Toronto, West Humber-Clairville has the most crime occurrence in 2023 with 1856 crime cases and also has the most population. The area with the second highest level of crime occurence is Downtown Younge East with 1259 cases. Additionally, the area that came third in terms of highest crime occurrence is York University Height with 1236 cases. The fourth area where the most crime occurred is the Younge-Bay Corridor with 1117 cases. 

```{r}
# 10 neighborhood with the most crime

sorted_table <- neighborhood_crime_clean %>%
  arrange(desc(TOTAL_CRIME_2023)) %>%
  slice_head(n = 10)


sorted_table |>
  summarise(total_crime = TOTAL_CRIME_2023,
            .by = AREA_NAME) |>
  kable()

```

```{r}
area <- neighborhood_crime_clean$AREA_NAME
max_assault <- which.max(neighborhood_crime_clean$ASSAULT_2023)
max_auto_theft <- which.max(neighborhood_crime_clean$AUTOTHEFT_2023)
max_bike_theft <- which.max(neighborhood_crime_clean$BIKETHEFT_2023)
max_break_enter <- which.max(neighborhood_crime_clean$BREAKENTER_2023)
max_homicide <- which.max(neighborhood_crime_clean$HOMICIDE_2023)
max_shooting <- which.max(neighborhood_crime_clean$SHOOTING_2023)
max_robbery <- which.max(neighborhood_crime_clean$ROBBERY_2023)
max_theft_mv <- which.max(neighborhood_crime_clean$THEFTFROMMV_2023)
max_theft_over <- which.max(neighborhood_crime_clean$THEFTOVER_2023)


type_of_crime <- data.frame(
  Type_of_Crime= c("Assault",  "Auto Theft", "Bike Theft", "Breaking and Entering",                "Robbery", "Homicide", "Shooting", "Theft from moving vehicles", "Theft Over"), Area_with_max = c(area[max_assault], area[max_auto_theft], area[max_bike_theft], area[max_break_enter], area[max_homicide], area[max_shooting], area[max_robbery], area[max_theft_mv], area[max_theft_over]))


kable(type_of_crime)

```

## 4. Discussions.

```{r}
#scatterplot
ggplot(neighborhood_crime_clean, aes(x = TOTAL_CRIME_2023, y = POPULATION_2023)) +
   geom_point() + 
  labs(title = "Correlation between crime and population density", x = "number of crimes", y = "Y-axis") +
    theme_minimal() + # Minimal theme
  theme(
    plot.title = element_text(size = 16),  # Title size and bold
    axis.title = element_text(size = 14),  # Axis title size
    axis.text = element_text(size = 12),  # Axis text size
    legend.position = "none",  # Remove legend
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank()   # Remove minor grid lines
  )
```

## 5. Conclusion.
