---
title: "Does denser population area has more crime than others?"
author: "Thu Dong"
format: pdf
editor: visual
abstract: "Toronto was reported to have an increasing crime rate in 2022 and is believed to keep rising in the future. Therefore, this research has investigated which area of Toronto has the highest level of crime occurrences and whether it is associated with the population density of the area. This paper reveals that areas with higher populations do have more crime. However, there is no causal relationship between crime level and population in the area."
bibliography: references.bib
---

```{r}
#| include: false
#install.packages("kableExtra")
#install.packages( "psych")
```

```{r}
#| include: false
#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(psych) 
library(here)
```

```{r}
#| include: false
neighborhood_crime_clean <-
  read_csv(
    here("/cloud/project/outputs/data/neighborhood_crime_2023_clean.csv"),
    show_col_types = FALSE
  )
```

## Table of Content

## 1. Introduction

Toronto is considered one of the safest cities in Canada and North America with homicide rates of approximately 1.95 per 100,000 people in 2020 (Armstrong and Jaffray 2021). However, it was reported that there was a 14% increase in the overall crime rate in 2022 and it will continue to increase in 2023 (Rosenbaum 2023). It is believed that more populated areas tend to have higher crime incidents. This paper attempts to investigate whether areas with higher populations experience higher levels of crime occurrences.  

There have been many studies in criminology that discuss the relationship between population size and crime level. One of the first works in criminology is research conducted across France in 1825 by Balbi and Guerry which reveals that the area with a higher population has more crime occur than others. However, it was later discussed that these 2 relationships are not linear and might be a correlation-based conclusion. (Friendly 2007). On the other hand, some other research in criminology disproved the theory that crime increases with population. (Chamlin and Cochran 2004). This research argued that there is no direct association between population and crime rate and other factors that are related to population growth such as quality of life have a stronger correlation to the volume of crime than population. Most of the research has used crime rates to measure the intensity of crime in an area, standardize a measure, and manage risk assessments. This paper will be using crime volume or the number of crimes to assess the magnitude of crime level. Using crime volume also helps policymakers to understand the safety state of each area to give efficient resource allocation. 

To discuss the relationship between population density and crime volume, this paper will present in the following section: Data, Results, Discussion, and Conclusion. In the Data section, I will discuss the nature of the data obtained through the City of Toronto's OpenDataToronto Library and the steps I took to clean the data. The Result section will highlight some trends found in the data. Additionally, the Discussion Section will explore the relationship between the level of crime occurrence and population density. Finally, The conclusion section will summarize the main findings of the paper. 

## 2. Data

The data used in this paper was obtained through the City of Toronto's OpenDataToronto Library. The data set is called 'Neighborhood Crime Rate' ("Open Data Dataset" 2024). Data was collected and analyzed using the statistical programming software R[@citeR] (R Core Team 2023 ), with additional support from tidyverse[\@wickham](Wickham%20et%20al.%202019), ggplot2 [@book] (Wickham 2016), dplyr [\@wh](Wickham%20et%20al.%202023), readr [@readr]\[(Wickham, Hester, and Bryan 2023), here[@here] (Müller K 2020), knitr [@knitr] ( Yihui Xie 2014), kableExtra [@kableE] ( Zhu H 2021), psych [@psych] ( William Revelle 2024).

### **2.1.Crime volume in 2023.**

The original "Neighborhood Crime Rate" Data set includes 185 variables and 158 observations. It records the number of Assaults, Auto Theft, Break and Entering, Robbery, Theft Over, Homicide, and Shooting from the year 2014 to the year 2023, as well as the crime rate of each crime in each year. 

To get the data for crime volume in 2023, I selected 8 columns: Assaults, Auto Theft, Breaking and Entering, Robbery, Theft Over, Homicide, and Shooting in 2023 since it is the closest data recorded. Additionally, all the "NA" values were set to 0 to allow for mathematical operation on the data.  Then, I summed the number of each crime to get the total crime of each area in the year 2023. Finally, I renamed all the columns needed to allow for clarity.

### **2.2. Population in 2023.**

Data for area in Toronto and population in 2023 was available in the " Neighborhood Crime rate" dataset. Therefore, the name of the area and the population in 2023 were added to the clean data set. shows the first 5 rows of the data set with area name, population, and the number of crime occurrences i

```{r}
#| echo: false
neighborhood_crime <- select(neighborhood_crime_clean, AREA_NAME, POPULATION_2023, TOTAL_CRIME_2023)

write_csv(neighborhood_crime, "neighborhood_crime_analysis.csv")

kable(head(neighborhood_crime,5), 
      col.names = c("Area in Toronto", "population in 2023", "total crime cases in 2023"),
      caption = " Sample of data with crime cases and population in Toronto areas in 2023  ") %>%
  kable_styling(full_width = FALSE)

```

## 3. Result

### **3.1. Toronto Area and the population in 2023.**

There are 158 areas divided in Toronto. Out of the 158 areas in Toronto,  Beechborough-Greenbrook is the part with the lowest population of 7,057 and West Humber-Clairville is the area with the highest population at 36,388. Table 2 shows the 10 most populated areas in Toronto. The second most populated area is Mount Olive-Silverstone-Jamestown with 35,386 people but the third highest populated area is the Annex with 34502 people. Other area that are highly populated are Glenfield-Jane Heights, Harbourfront-CityPlace, Agincourt North, etc... 

```{r}
#| echo: false
#Summary table
summary_result <- describe(neighborhood_crime$POPULATION_2023)
#summary_result

#area with lowest population
min_populated_area <- neighborhood_crime$AREA_NAME[which.min(neighborhood_crime$POPULATION_2023)]

sorted_population_table <- neighborhood_crime_clean %>%
  arrange(desc(POPULATION_2023)) %>%
  slice_head(n = 10)


sorted_population_table |>
  summarise(population = POPULATION_2023,
            .by = AREA_NAME) |>
  kable(col.names = c("Area in Toronto", "population in 2023"), 
        caption = "Top 10 areas with highest population in Toronnto")

```

### **3.2. Toronto Area and crime volume.**

Table 3 shows 10 areas with the highest level of criminal incidents in Toronto. Out of the 158 areas in Toronto, West Humber-Clairville has the most crime occurrence in 2023 with 1856 crime cases and also has the most population. The area with the second highest level of crime occurence is Downtown Younge East with 1259 cases. Additionally, the area that came third in terms of highest crime occurrence is York University Height with 1236 cases. The fourth area where the most crime occurred is the Younge-Bay Corridor with 1117 cases. 

```{r}
#| echo: false
# 10 neighborhood with the most crime

sorted_table <- neighborhood_crime_clean %>%
  arrange(desc(TOTAL_CRIME_2023)) %>%
  slice_head(n = 10)


sorted_table |>
  summarise(total_crime = TOTAL_CRIME_2023,
            .by = AREA_NAME) |>
  kable(   
    col.names = c("Area in Toronto", "number of crime in 2023"),
    caption = "Top 10 areas with the most crime in Toronto ")

```

Table 4 illustrates which area in Toronto has the highest crime occurrence of a particular type of crime. Not only did West Humber-Clairville have the most crime cases, but it also was the area that had the most auto theft incidences, theft from moving vehicle incidences, and theft over incidences. While Downtown Yonge East came in second place in terms of the highest crime level, it has the highest occurrences of assaults. York University Height was reported to have the most shooting cases and breaking and entering cases. Additionally, the area with the highest bike theft cases is Yonge- Bay Corridor. Finally, Moss Park is the area with the most robbery cases and Mount Olive-Silverstone-Jamestown is recorded to have the highest homicide cases. 

```{r}
#| echo: false
area <- neighborhood_crime_clean$AREA_NAME
max_assault <- which.max(neighborhood_crime_clean$ASSAULT_2023)
max_auto_theft <- which.max(neighborhood_crime_clean$AUTOTHEFT_2023)
max_bike_theft <- which.max(neighborhood_crime_clean$BIKETHEFT_2023)
max_break_enter <- which.max(neighborhood_crime_clean$BREAKENTER_2023)
max_homicide <- which.max(neighborhood_crime_clean$HOMICIDE_2023)
max_shooting <- which.max(neighborhood_crime_clean$SHOOTING_2023)
max_robbery <- which.max(neighborhood_crime_clean$ROBBERY_2023)
max_theft_mv <- which.max(neighborhood_crime_clean$THEFTFROMMV_2023)
max_theft_over <- which.max(neighborhood_crime_clean$THEFTOVER_2023)


type_of_crime <- data.frame(
  Type_of_Crime= c("Assault",  "Auto Theft", "Bike Theft", "Breaking and Entering","Robbery", "Homicide", "Shooting", "Theft from moving vehicles", "Theft Over"), 
  Area_with_max = c(area[max_assault], area[max_auto_theft], area[max_bike_theft], area[max_break_enter], area[max_homicide], area[max_shooting], area[max_robbery], area[max_theft_mv], area[max_theft_over]))


kable(type_of_crime, col.names = c("Type of crime", "Area in Toronto"),
    caption = " Area with the highest level of a particular crime.")

```

## 4. Discussions.

Overall, the data reveals that there is a clear relationship between the number of crime occurrences and the population in specific areas, as displayed in Figure 1. It is shown that the increase in population is also associated with the increase in the level of criminal incidences. This positive correlation is moderately strong given that the data shows a relatively tightly clustered trend. There is an outlier which is believed to be West Humber-Clairville which has a population of 36388 and 1856 criminal cases. Additionally, there has been a debate on whether or not the relationship between crime level and population is linear. The data demonstrates that even though there is a clear positive relationship between the number of crime cases and population in specific areas, there could be a nonlinear relationship between the 2 variables. 

```{r}
#| echo: false
#scatterplot
ggplot(neighborhood_crime_clean, aes(x = POPULATION_2023, y = TOTAL_CRIME_2023)) +
   geom_point() + 
  labs(title = "Correlation between crime and population density", x = "Population in 2023", y = "number of cases in 2023") +
    theme_minimal() + # Minimal theme
  theme(
    plot.title = element_text(size = 13),  # Title size and bold
    axis.title = element_text(size = 10),  # Axis title size
    axis.text = element_text(size = 10),  # Axis text size
    legend.position = "none",  # Remove legend
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank()   # Remove minor grid lines
  )
```

There are some limitations regarding the validity of this analysis. Firstly, Some threats to the internal validity of the analysis are wrong functional form and missing data. Since the scatterplot only shows a positive correlation between crime incidences and population in an area, this is an association-based paper and not a causal-based research. Additionally, as previously discussed, there has been some research that the relationship between the area's population and level of crime is nonlinear. The research could also be biased because of some unavailable data noted as "NA" in the data set which has been transformed into "0" for mathematical operation purposes. Secondly, there may be some threat to the external validity of this research. Since the population of interest in this research is crime and population in Toronto in 2023, the same association can't be drawn using a different data set without further investigation. 

## 5. Conclusion.

This paper investigates which area in Toronto has a higher population and which has more crime as well as the relationship between crime level and the area's population. The outcome revealed that there is a positive relationship between the number of crime cases and the population in a particular area. This is a correlation research so there is no conclusion on the causal inferences of the two variables. Going onwards, this research will establish the causal inference on whether the increase in population leads to the increase in crime by including some other omitted variables such as income, quality of life, etc. 
